Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-07 paper/paper.tex
I don't want my office room number there :-)
fixing figures

\author[0000-0001-5725-9329]{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0002-3962-9274]{Boris Leistedt}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{NASA Einstein Fellow}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it is a generative model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

Here we present a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallax measurements, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

@@

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.png}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper-left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper-right and lower-left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower-left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the expectation value of the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower-right panel shows the expectation value, and $1\sigma$ uncertainties, of the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value. The blue points are better than the black for doing stuff.}
\label{fig:toy}
\end{figure}

In this paper we use the noise-deconvolved \cmd\ as a prior to infer more precise parallax probability distribution functions (PDFs) from \gaia\ measurements.
Understanding the validity of this method can be challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we demonstrate the general methodology using a
simpler toy model with simulated data.
Instead of inferring parallax PDFs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer posterior PDFs of the value $Y$.
We perform this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = m\,x + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness or intrinsic scatter of the \cmd\ due to age, mass, and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in \figurename~\ref{fig:toy}, upper-left panel.
Similarly, noisy samples from this distribution are shown in the upper-right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample from the distribution, from $y_n$, a sample from the noisy distribution.

@@

Under these assumptions, we would like to infer the true parallax, $\varpi_{\true}$, for each star given its measured parallax, $\varpi$, and associated uncertainty, $\sigma_{\varpi}$, from \tgas.
A first application of Bayes theorem leads to
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi})$ is the likelihood of the observed parallax, $p(\varpi_{\true})$ is a prior PDF for the true
parallax, and $Z$ is a normalization constant (the model evidence, or marginalized likelihood).
The strength of our inference lies in how we generate our prior.
Instead of assuming a functional form for the spatial distribution of stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar models (\citealt{gaia16}), we build a noise-deconvolved estimate of the \cmd\ from the observed colors and absolute magnitudes of our full set of stars.
This is because supplementary to the trigonometric measure of distance, the photometry of stars is also informative of their distances.
The intrinsic temperatures and luminosities of stars tightly correlate, leading to the observed colors and absolute magnitudes of stars also tightly correlating, usually visualized as the \cmd.
The tight correlation of the \cmd\ benefits our inference because the luminosity or absolute magnitude of a star $M$, when compared with its brightness or apparent magnitude $m$, contains distance information $d$ via the fundamental relation $m=M+5\log(d/10\, \mathrm{pc})$.
We include this in the inference by using the \cmd\ as part of our prior for $\varpi_{\true}$.
Specifically, we use the \cmd\ generated from the data themselves, as detailed in the next section.

@@

We use stars crossmatched in \tgas\ and \tmass.
The match was done using a nearest-neighbor algorithm with a search radius of $4"$.\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with it's associated error bars, calculated via subsampling.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

To generate the prior and evaluate the likelihood, we need to
correct the \tmass\ photometry for dust extinction.
With the emergence of 3D dust maps, it is now possible to estimate dust
corrections to stars within the Milky Way. The challenge is getting a
proper estimate of the distance to the star before inferring
$\varpi_{\true}$, since we need a dust correction prior to doing our
inference. Dust corrections are particularly non trivial for stars
with poor parallax signal to noise; these stars have
a likelihood that is consistent with infinite distance and can
therefore have severe dust corrections. This is most obvious in the
giant stars, which tend to have the lowest signal to noise, and can
therefore have dust corrections $> 2$ magnitudes. Although the dust
model is probabilistic, as well as our distance inference, to move
forward, we break from a fully Bayesian framework in which we might
infer the distance and the dust simultaneously (more about this in the
discussion). Instead, we correct for dust using a point estimate from
the 3D dust map \citep{green15}. We determine this dust correction by
iteratively inferring, and then sampling, our posterior PDF over
$\varpi_{\true}$. We first generate our prior using the observed,
attenuated photometry. Using this raw prior, we infer more precise
parallaxes to all the stars in our sample. We use this more
precise parallax posterior to get a measurement of dust in the 3D dust
map for each star. We then iterate this process 10 times, however
the dust values seem to converge after about 2-3 iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.

In detail, we take the $5\%$
quantile of each parallax posterior (the closest part) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the likelihood
function into an unphysical, extremely blue part of the \cmd\ and thus bias the
parallax inference. At the 3D position of each star, we sample the \texttt{mode=median} value of the probabilistic dust map.
We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K_s]$ respectively (see Table 6 in
\citealt{schlafly11}).
The dust values we obtain in our final
iteration, and which we apply to the photometry in our inference, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; this leads to a less severe deconvolution, as mentioned previously.
In other words, the features of our inferred
  \cmd\ are conservatively broad.\footnote{This can be understood by considering the following procedure: if one incrementally increased the noise levels (leaving the data unchanged; just increasing the declared uncertainties), the deconvolved distribution would be increasingly narrower, since it targets the distribution that would generate the noiseless observations, as well as the observed data once noise is added. Thus, under-estimating the uncertainties leads to a conservative under-deconvolution.}
We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross-matched catalog. The large missing areas are due to the \psone\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the y-axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result which we illustrated in the toy model above: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog. This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
And $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, or have a fraction decrease in variance by a factor of $>4$, shown as the other dotted line at $y=0.14$.

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1-4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that distance estimates directly derived from the inverse of the observed (noisy) parallax are unreliable (significantly biased and uncertain) when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.
With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more clustered and sharply peaked around the ``true" distance to M67 than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age variation on the main sequence mixed
with a Malmquist bias of seeing a much larger volume of upper main
sequence stars \citep{malmquist22}. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a Malmquist
bias.

@@

  \cmd\ space. Thus, this has a small effect on our results, and could be improved by modelling the noise distribution simultaneously with the \cmd.

@@

  the \tgas--\tmass\ intersection, the model \emph{is} representative of

@@

  A similar argument can be made for \textit{internal} selection effects, such as inhomogeneities of the depth and noise distributions, which we have neglected. 

@@

This research was partially supported by
  the \acronym{NSF} (\acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  and the Moore-Sloan Data Science Environment at \acronym{NYU}. 
  BL was supported by the National Aeronautics and Space Administration through Einstein Postdoctoral Fellowship Award Number PF6-170154.
It made use of the \acronym{NASA} Astrophysics Data System.
All the code used in this project is available online\footnote{\url{https://github.com/andersdot/photoParallax}} under an open-source license.

>>>>

\author[0000-0001-5725-9329]{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0002-3962-9274]{Boris Leistedt}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise (and beautiful) because it is a generative model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

Here we deliver a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallax measurements, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

@@

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.pdf}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper-left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper-right and lower-left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower-left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the expectation value of the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower-right panel shows the expectation value, and $1\sigma$ uncertainties, of the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value. The blue points are better than the black for doing stuff.}
\label{fig:toy}
\end{figure}

In this paper we use the noise-deconvolved \cmd\ as a prior to infer more precise parallax probability distribution functions (PDFs) from \gaia\ measurements.
Understanding the validity of this method can be challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we demonstrate the general methodology using a
simpler toy model with simulated data.
Instead of inferring parallax PDFs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer posterior PDFs of the value $y$.
We perform this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the $y$ direction, with a running mean $\mu = m\,x + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness or intrinsic scatter of the \cmd\ due to age, mass, and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in \figurename~\ref{fig:toy}, upper-left panel.
Similarly, noisy samples from this distribution are shown in the upper-right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample from the distribution, from $y_n$, a sample from the noisy distribution.

@@

Under these assumptions, we would like to infer the true parallax, $\varpi_{\true}$, for each star given its measured parallax, $\varpi$, and associated uncertainty, $\sigma_{\varpi}$, from \tgas.
A first application of Bayes theorem leads to
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi})$ is the likelihood of the observed parallax, $p(\varpi_{\true})$ is a prior PDF for the true
parallax, and $Z$ is a normalization constant (the model evidence, or marginalized likelihood).
The strength of our inference lies in how we generate our prior.
Instead of assuming a functional form for the spatial distribution of stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar models (\citealt{gaia16}), we build a noise-deconvolved estimate of the \cmd\ from the observed colors and absolute magnitudes of our full set of stars.
This is because supplementary to the trigonometric measure of distance, the photometry of stars is also informative of their distances.
The intrinsic temperatures and luminosities of stars tightly correlate leading to the observed colors and absolute magnitudes of stars also tightly correlating, usually visualized as the \cmd.
The tight correlation of the \cmd\ benefits our inference because the luminosity or absolute magnitude of a star $M$, when compared with its brightness or apparent magnitude $m$, contains distance information $d$ via the fundamental relation $m=M+5\log(d/10\, \mathrm{pc})$.
We include this in the inference by using the \cmd\ as part of our prior for $\varpi_{\true}$.
Specifically, we use the \cmd\ generated from the data themselves, as detailed in the next section.

@@

We use stars crossmatched in \tgas\ and \tmass\.
The match was done using a nearest-neighbor algorithm with a search radius of $4"$.
\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with it's associated error bars.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

To generate the prior and evaluate the likelihood, we need to
correct the \tmass\ photometry for dust extinction.
With the emergence of 3D dust maps, it is now possible to estimate dust
corrections to stars within the Milky Way. The challenge is getting a
proper estimate of the distance to the star before inferring
$\varpi_{\true}$, since we need a dust correction prior to doing our
inference. Dust corrections are particularly non trivial for stars
with poor parallax signal to noise; these stars have
a likelihood that is consistent with infinite distance and can
therefore have severe dust corrections. This is most obvious in the
giant stars, which tend to have the lowest signal to noise, and can
therefore have dust corrections $> 2$~mag. Although the dust
model is probabilistic, as well as our distance inference, to move
forward, we break from a fully Bayesian framework in which we might
infer the distance and the dust simultaneously (more about this in the
discussion). Instead, we correct for dust using a point estimate from
the 3D dust map \citep{green15}. We determine this dust correction by
iteratively inferring, and then sampling, our posterior PDF over
$\varpi_{\true}$. We first generate our prior using the observed,
attenuated photometry. Using this raw prior, we infer more precise
parallaxes to all the stars in our sample. We use this more
precise parallax posterior to get a measurement of dust in the 3D dust
map for each star. We then iterate this process ten times, however
the dust values seem to converge after a few iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.

In detail, we take the $5\%$
quantile of each parallax posterior (the closest part) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the likelihood
function into an unphysical, extremely blue part of the \cmd\ and thus bias the
parallax inference. At the 3D position of each star, we sample the \texttt{mode=median} value of the probabilistic dust map.
We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K_s]$ respectively (see Table 6 in
\citealt{schlafly11}).
The dust values we obtain in our final
iteration, and which we apply to the photometry in our inference, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; this leads to a less severe deconvolution, as mentioned previously.
In other words, the features of our inferred
  \cmd\ are conservatively broad (broader than if we over-estimated
  the errors which would have lead to excessive deconvolution and
  unrealistically narrow features). We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross-matched catalog. The large missing areas are due to the \psone\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the $y$ axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog. This more directly shows what was alluded to in Figure~ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
And $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, or have a fraction decrease in variance by a factor of $>4$, shown as the other dotted line at $y=0.14$.

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1 through 4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$ to $900~{\rm pc}$ with an estimated age of $3$ to $5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.
With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the "true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more clustered and sharply peaked around the "true" distance to M67 than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age variation on the main sequence mixed
with a Malmquist bias of seeing a much larger volume of upper main
sequence stars. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a Malmquist
bias (\citealt{malmquist22}).

@@

  \cmd\ space.

@@

  the \tgas--\tmass\ intersection, the model \emph{is} representive of

@@

@@

This project was partially supported by This research was partially supported by
  the \acronym{NSF} (\acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  and the Moore-Sloan Data Science Environment at \acronym{NYU}.
It made use of the \acronym{NASA} Astrophysics Data System.
All the code used in this project is available online\footnote{\url{https://github.com/andersdot/photoParallax}} under an open-source license.
